Inference using context-dependent models of mutation via bounds on information propagation

### 2021-12-09 writeups/context-dependence.tex
further work

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{bm}
\usepackage{todonotes}

@@

% Model-based inference neighborhood dependence
% Nearby sites affecting the likelihood
% Rigorous inference with neighborhood dependency
% Statistical inference with local/neighborhood interdependency
% It depends on the context:
%

\title{Inference using context-dependent models of mutation via bounds on information propagation}
\author{}

@@

\begin{abstract}
    Although the rates at which positions in the genome mutate
    are known to depend not only on the nucleotide to be mutated
    but also on neighboring nucleotides,
    it remains challenging to do phylogenetic inference
    using models of context-dependent mutation.
    In these models,
    the effects of one mutation may in principle propagate
    to faraway locations,
    making it difficult to compute exactly likelihoods.
    This paper shows how to use bounds on the propagation of dependency
    to compute likelihoods of mutation of a given segment of genome
    by marginalizing over sufficiently long flanking sequence.
    This can be used for maximum likelihood or Bayesian inference.
    Protocols examining residuals and iterative model refinement
    are also discussed.
    Tools for efficiently working with these models
    are provided in an R package,
    that could be used in other applications.
    The method is used to examine context dependence of mutations
    since the common ancestor of humans and chimpanzee.
\end{abstract}

@@

\begin{figure}
    \begin{center}
        \includegraphics{overview}
    \end{center}
    \caption{
        An overview of the method for a neighbor-dependent model.
        The red trapezoidal region represents the ``range of influence'',
        i.e., possible locations and times at which a mutations is likely
        to affect the mutation outcome at
        the sites of interest (the three sites in the center of the final sequence).
        The blue T shape is called a \emph{T-mer},
        which is how our approximation to the process is parameterized.
        The central, smaller segment of the T is called the \emph{base} of the T-mer,
        and the key building block computes the transition probability
        for this subsequence
        conditioned on the ancestral state at the \emph{overhang}
        of the top of the T.
        \label{fig:overview}
    }
\end{figure}

@@

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that will be sufficient to compute the full likelihood,
at least arbitrarily.
%EM do we really want x and y here, which I've been thinking of as being the full sequence?
%EM Note we'll need to change a and b below.
%PR I think this is okay? x and y are not exactly the same here, but are still sequences?
%PR To me it seems less confusing than introducing another variable?
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
(see Figure \ref{Tmer-dependency}).
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
We emphasize that the overhang of the T-mers is typically bigger than the
context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).

@@

The approximations in~\eqref{eqn:window_approx} and~\eqref{eqn:full_approx} hold
if the overhang, $\ell$, is long enough.
We present intuition about these formulas here, which provides a way to pick the overhang lengths $\ell$ and $r$;
a precise statement is given in Appendix~\ref{ss:approx_pf}.
Suppose that each instantaneous change depends on positions at most $w$ sites away,
and refer to $w$ as the ``window width''.
Then, the mutation process at any particular site
depends on the initial sequence at that site and within distance $w$ on either side\ldots
\emph{and} the outcomes of any mutations that have occurred within that window
(see Figure \ref{Tmer-dependency}).

@@

For instance, in the CpG model the window width is $w=1$
because the process only depends on neighboring sites;
and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
then taking $\ell = r = 3$ allows us to compute T-mer probabilities to within an error of 0.007,
since this is the probability that a Poisson with mean 0.75 will be larger than 3.
If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.

@@

Instead, we will compute the probability that
we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$,
summing over possible (long) patterns at the root.
Concretely: pick a random location $I$ in the sequence
and let $X_{I-\ell}^{(n)}(0) = \rho_I$ be the (long) pattern seen there at the root.
Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho_I=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho_I$ is proportional to
% \[
%     \P\{ \rho_I=z \mid X_{I-\ell}^{(n)}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(\ell)}(t_Y)=y \}
    =
    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
\end{align}
Denote this quantity $p_{n,\ell,r}(x, y)$.
This can be computed without much more effort than the simpler case above,
given the frequencies at the root, as described in greater generality below.
These calculations can be done using the same sparse matrix methods as above.
For instance, to compute
\[
    \sum_{z \in \calS^{n}}
        \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
    =
    \sum_{z \in \calS^{n}}
    \left( e^{t_X G(n)} \right)_{zx}
    \pi(z)
    \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_Y G(n)} \right)_{zw} U_{wy}$ as above,
multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $e^{t_X G(n)}$.

We might approximate the full likelihood as follows,
assuming for the sake of notational simplicity that we have a few positions
of $X$ observed before the start and after then end of the sequence
(so that, e.g., we can refer to $x_{n+1}, \ldots, x_{n+r}$):
\begin{align*}
    \like(x, y)
    &=
    \prod_{i=1}^L
        \P\{ X_{i+r} = x_{i+r},\; Y_i = y_i
            \given X_{<i+r} = x_{<i+r},\; Y_{<i} = y_{<i} \} \\
    &\approx
    \prod_{i=1}^L
        \P\{ X_{i+r} = x_{i+r},\; Y_i = y_i
            \given X_{i-\ell}^{(n-1)} = x_{i-\ell}^{(n-1)},\; Y_{i}^{(m-1)} = y_{i}^{(m-1)} \} \\
    &=
    \prod_{i=1}^L
    \frac{
        p_{n,\ell,r}\left( x_{i-\ell}^{(n)}, y_i^{(m)} \right)
    }{
        p_{n-1,\ell,r}\left( x_{i-\ell}^{(n-1)}, y_i^{(m-1)} \right)
    }
\end{align*}
We expect this to be a good approximation to the true likelihood
when the same conditions above are met, taking $t$ to the be the largest distance
between two taxa in the tree.

%EM I think this should be pruning, not peeling.
%PR I'll take your word for it! Wikipedia suggests either would do?
%EM Want to check the modifications made above before continuing those to here.
\subsection{Phylogenetic pruning}

@@

\begin{figure}
    \begin{center}
    \includegraphics{pruning-schematic}
    \end{center}
    \caption{
        A depiction of the four steps in computing the likelihood
        of finding particular ``short'' subsequences at $Y$ and $Z$
        given the ``long'' subsequence at $X$: see text for details.
        \label{fig:pruning}
    }
\end{figure}

To compute the likelihood on a more complicated tree, consider the following,
with the three-taxon tree of figure \ref{fig:pruning} as an example,
where we are counting occurrences of (long) $n$-tuples in taxon $X$,
and (shorter) $m$-tuples in taxa $Y$ and $Z$.
The likelihood in this example is found by summing over the root state (the summation over $u$)
and the interior node ($v$):
\begin{align} \label{eqn:three_taxa_likelihood}
  \begin{split}
    & \P\{X_{I-\ell}^{n}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
    &\qquad
      =
      \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell,r}(t_C,u,z)
      \sum_{v \in \calS^{n}} p_{n,0,0}(t_{AB},u,v) p_{n,0,0}(t_{A},v,x) p_{m,\ell,r}(t_B,v,y)
  \end{split}
\end{align}
The steps for computing this are then:
\begin{enumerate}

@@

In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
However, we can reduce dimensionality by replacing the last step with
\begin{enumerate}

@@

We constructed models from three, nested, sets of mutation motifs.
All models are strand-symmetric.
\begin{itemize}
    \item[(basic)]
        Single-base changes, with separate rates for transitions ($\mu_v$),
        \nA$\leftrightarrow$\nT{} ($\mu_{AT}$), \nC$\leftrightarrow$\nG{} ($\mu_{CG}$),
        and other transversions ($\mu_{r}$):
          \begin{center}
            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nA  &   \nT   &  $\mu_{AT}$ & \nG  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
                \nC  &   \nG   &  $\mu_{CG}$ & \nC  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nC   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
            \end{tabular}
          \end{center}

@@

@@

\begin{table}
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\
          \hline
          \nA  &   \nT           & 0.10 & 0.0996 \\
          \nT  &   \nA           & 0.10 & 0.0982 \\
          \nC  &   \nG           & 0.15 & 0.1505 \\
          \nG  &   \nC           & 0.15 & 0.1497 \\
          \nA  &   \nC           & 0.08 & 0.0784 \\
          \nT  &   \nG           & 0.08 & 0.0792 \\
          \nA  &   \nG           & 0.08 & 0.0798 \\
          \nT  &   \nC           & 0.08 & 0.0795 \\
          \nC  &   \nA           & 0.12 & 0.1202 \\
          \nG  &   \nT           & 0.12 & 0.1196 \\
          \nC  &   \nT           & 0.12 & 0.1198 \\
          \nG  &   \nA           & 0.12 & 0.1193 \\
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
           \hline
        \end{tabular}
  \end{center}
  \caption{ Results from iterative fitting to a $10^6$bp sequence
    evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
    were constrained to be equal (DNA strand symmetry),
    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
    \label{tab:cpg_results} }
\end{table}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors}
    \end{center}
    \caption{
        Estimated average mutation rates
        for the motifs in the ``Hominid data'' section
        since the common ancestor of humans and chimpanzee.
        Rates are scaled so that the mean time to common ancestor is 0.001 units --
        if we take this to be 6 million years \citep{scally2012insights,langergraber2012generation}
        then rates are in units of mutations per 6,000 years.
        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail,
        and figure \ref{sfig:biochem_results_relative} shows the rates between categories relative to each other.
        \label{fig:biochem_results}}
\end{figure}

@@

The methods described here provide a way to efficiently calculate full likelihoods,
which in many circumstances provide the optimal basis for inference \citep{neyman1933problem},
fast enough to allow Bayesian inference from Monte Carlo sampling.
The method is fast thanks to its dependence only only local neighborhoods
(so the time does not scale with sequence length once T-mer abundances are counted)
and its reliance on sparse matrices whose structure can be precomputed and cached.
It avoids approximations made by previous methods thanks to the use of T-mers
and explicit quantification of the necessary length of the ``arms'' of the T-mers.
We first discuss possible uses of the method,
followed by challenges (and possible solutions).

@@

\subsection{Acknowledgements}
The initial impetus for this project came from Matt Dean;
along the way substantial encouragement and useful suggestions were provided by
Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
Many thanks are also due Jessica Crisci for curating the human-chimp data.
The majority of this work was done while PR was at USC.

%EM TODO grant acknowledgements

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{bm}
\usepackage{todonotes}
\usepackage{authblk}
\renewcommand\Affilfont{\itshape\small}

@@

% TODO include github information

%EM I'm going to suggest "dependence propagation" rather than information
\title{Inference using context-dependent models of mutation via bounds on information propagation}
\author[1]{Peter L Ralph}
\author[2,3,4,5]{Frederick A Matsen IV}
\author[1,*]{Peter L Ralph}

\affil[1]{Institute of Ecology \& Evolution, Department of Biology, University of Oregon}
\affil[2]{Computational Biology Program, Fred Hutchinson Cancer Research Center, United States}
\affil[3]{Department of Genome Sciences, University of Washington}
\affil[4]{Department of Statistics, University of Washington}
\affil[5]{Howard Hughes Medical Institute}
\affil[*]{Corresponding author: plr@uoregon.edu}

@@

\begin{abstract}
    Although the rates at which positions in the genome mutate
    are known to depend not only on the nucleotide to be mutated,
    but also on neighboring nucleotides,
    it remains challenging to do phylogenetic inference
    using models of context-dependent mutation.
    In these models,
    the effects of one mutation may in principle propagate
    to faraway locations,
    making it difficult to compute exact likelihoods.
    This paper shows how to use bounds on the propagation of dependency
    to compute likelihoods of mutation of a given segment of genome
    by marginalizing over sufficiently long flanking sequence.
    This can be used for maximum likelihood or Bayesian inference.
    Protocols examining residuals and iterative model refinement
    are also discussed.
    Tools for efficiently working with these models
    are provided in an R package,
    that could be used in other applications.
    The method is used to examine context dependence of mutations
    since the common ancestor of humans and chimpanzee.
\end{abstract}

@@

\begin{figure}
    \begin{center}
        \includegraphics{overview}
    \end{center}
    \caption{
        An overview of the method for a neighbor-dependent model, showing an
        initial sequence on top, a collection of mutations progressing through
        time, and a final sequence.
        The pink trapezoidal region represents the ``range of influence'',
        i.e., possible locations and times at which a mutations is likely
        to affect the mutation outcome at
        the sites of interest (the three sites in the center of the final sequence).
        The blue T shape is called a \emph{T-mer},
        which is how our approximation to the process is parameterized.
        The central, smaller segment of the T is called the \emph{base} of the T-mer,
        and the key building block computes the transition probability
        for this subsequence
        conditioned on the ancestral state at the additional \emph{overhang}
        sites of the top of the T.
        \label{fig:overview}
    }
\end{figure}

@@

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that will be sufficient to compute the full likelihood,
at least arbitrarily.
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites
(see Figure \ref{fig:Tmers}).
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
We emphasize that the overhang of the T-mers is typically bigger than the
context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).

@@

The approximations in~\eqref{eqn:window_approx} and~\eqref{eqn:full_approx} hold
if the overhang, $\ell$, is long enough.
We present intuition about these formulas here, which provides a way to pick the overhang lengths $\ell$ and $r$;
a precise statement is given in Appendix~\ref{ss:approx_pf}.
Suppose that each instantaneous change depends on positions at most $w$ sites away,
and refer to $w$ as the ``window width''.
Then, the mutation process at any particular site
depends on the initial sequence at that site and within distance $w$ on either side\ldots
\emph{and} the outcomes of any mutations that have occurred within that window
(see Figure \ref{fig:Tmers}).

@@

For instance, in the CpG model the window width is $w=1$
because the process only depends on neighboring sites;
and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
%EM suggest "every site will have 75\% chance of changing", which is a little different than what you say because of multiple hits, right?
If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
then taking $\ell = r = 3$ allows us to compute T-mer probabilities to within an error of 0.007,
since this is the probability that a Poisson with mean 0.75 will be larger than 3.
If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.

@@

Instead, we will compute the probability that
we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$,
summing over possible (long) patterns at the root.
Concretely: pick a random location $I$ in the sequence
and let $X_{I-\ell}^{(n)}(0) = \rho_I$ be the (long) pattern seen there at the root.
Write $\pi(z)$ for the frequency of a given pattern $z$ seen there, i.e., $\P\{\rho_I=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho_I$ is proportional to
% \[
%     \P\{ \rho_I=z \mid X_{I-\ell}^{(n)}(t_X)=x\} \propto \pi(z) p_{n}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    %EM I think the superscript of Y_I should be (m)
    \P\{X_{I-\ell}^{(n)}(t_X)=x \text{ and } Y_I^{(\ell)}(t_Y)=y \}
    =
    \sum_{z \in \calS^{n}} \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y) .
\end{align}
Denote this quantity $p_{n,\ell,r}(x, y)$.
This can be computed without much more effort than the simpler case above,
given the frequencies at the root, as described in greater generality below.
These calculations can be done using the same sparse matrix methods as above.
For instance, to compute
\[
    \sum_{z \in \calS^{n}}
        \pi(z) \, p_{n,0,0}(t_X,z,x) \, p_{n,\ell,r}(t_Y,z,y)
    =
    \sum_{z \in \calS^{n}}
    \left( e^{t_X G(n)} \right)_{zx}
    \pi(z)
    \sum_{w \in \calS^{n}} \left( e^{t_Y G(n)} \right)_{zw} U_{wy} ,
\]
we can first compute $\sum_w \left( e^{t_Y G(n)} \right)_{zw} U_{wy}$ as above,
multiply rows by $\pi(z)$, and then matrix multiply by the transpose of $e^{t_X G(n)}$.

We might approximate the full likelihood as follows,
%EM In the previous full likelihood section, we discuss choosing l, r, and n, and use those in the equations.
%EM This makes the equations look a little different. Do we want to just clarify that we are writing the equations for general l, r, n, and that we refer the reader to the previous choice to choose these quantities?
assuming for the sake of notational simplicity that we have a few positions
of $X$ observed before the start and after then end of the sequence
(so that, e.g., we can refer to $x_{n+1}, \ldots, x_{n+r}$):
\begin{align*}
    \like(x, y)
    &=
    \prod_{i=1}^L
        \P\{ X_{i+r} = x_{i+r},\; Y_i = y_i
            \given X_{<i+r} = x_{<i+r},\; Y_{<i} = y_{<i} \} \\
    &\approx
    \prod_{i=1}^L
        \P\{ X_{i+r} = x_{i+r},\; Y_i = y_i
            \given X_{i-\ell}^{(n-1)} = x_{i-\ell}^{(n-1)},\; Y_{i}^{(m-1)} = y_{i}^{(m-1)} \} \\
    &=
    \prod_{i=1}^L
    \frac{
        p_{n,\ell,r}\left( x_{i-\ell}^{(n)}, y_i^{(m)} \right)
    }{
        p_{n-1,\ell,r}\left( x_{i-\ell}^{(n-1)}, y_i^{(m-1)} \right)
    }
\end{align*}
We expect this to be a good approximation to the true likelihood
when the same conditions above are met, taking $t$ to the be the largest distance
between two taxa in the tree.

\subsection{Phylogenetic pruning}

@@

\begin{figure}
    \begin{center}
    %EM We need to put these superscripts in parens, no? Also, should we use n instead of x+2l?
    \includegraphics{pruning-schematic}
    \end{center}
    \caption{
        A depiction of the four steps in computing the likelihood
        of finding particular ``short'' subsequences at $Y$ and $Z$
        given the ``long'' subsequence at $X$: see text for details.
        \label{fig:pruning}
    }
\end{figure}

To compute the likelihood on a more complicated tree, consider the following,
with the three-taxon tree of Figure~\ref{fig:pruning} as an example,
where we are counting occurrences of (long) $n$-tuples in taxon $X$,
and (shorter) $m$-tuples in taxa $Y$ and $Z$.
The likelihood in this example is found by summing over the root state (the summation over $u$)
and the interior node ($v$):
\begin{align} \label{eqn:three_taxa_likelihood}
  \begin{split}
    & \P\{X_{I-\ell}^{n}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
    &\qquad
      =
      %EM I think this should be n, ell, r, not m, ell, r
      \sum_{u \in \calS^{n}} \pi(u) p_{m,\ell,r}(t_C,u,z)
      \sum_{v \in \calS^{n}} p_{n,0,0}(t_{AB},u,v) p_{n,0,0}(t_{A},v,x) p_{m,\ell,r}(t_B,v,y)
  \end{split}
\end{align}
The steps for computing this are then:
\begin{enumerate}

@@

%EM I'm not really following this section. If we are only calculating likelihood for places where y and z are the same, why is phylogenetic inference useful?
In total there are $|\calS|^{3m+2\ell}$ possible data combinations.
However, we can reduce dimensionality by replacing the last step with
\begin{enumerate}

@@

We constructed models from three, nested, sets of mutation motifs.
All models are strand-symmetric.
\begin{itemize}
    \item[(basic)]
        Single-base changes, with separate rates for transitions ($\mu_v$),
        \nA$\leftrightarrow$\nT{} ($\mu_{AT}$), \nC$\leftrightarrow$\nG{} ($\mu_{CG}$),
        and other transversions ($\mu_{r}$):
          \begin{center}
            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
                %EM I think there were some typos in this table that I fixed.
                \nT  &   \nA   &  $\mu_{AT}$ & \nA  &   \nC   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nA  &   \nT   &  $\mu_{AT}$ & \nC  &   \nA   &  $\mu_{r}$ & \nG  &   \nA   &  $\mu_{v}$ \\
                \nC  &   \nG   &  $\mu_{CG}$ & \nG  &   \nT   &  $\mu_{r}$ & \nT  &   \nC   &  $\mu_{v}$ \\
                \nG  &   \nC   &  $\mu_{CG}$ & \nT  &   \nG   &  $\mu_{r}$ & \nC  &   \nT   &  $\mu_{v}$ \\
            \end{tabular}
          \end{center}

@@

%EM In my compilation there is a float that interrupts this flow.

@@

\begin{table}
  \begin{center}
        \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }ll}
          \hline
            x & y & truth & MLE \\
          \hline
          \nA  &   \nT           & 0.10 & 0.0996 \\
          \nT  &   \nA           & 0.10 & 0.0982 \\
          \nC  &   \nG           & 0.15 & 0.1505 \\
          \nG  &   \nC           & 0.15 & 0.1497 \\
          \nA  &   \nC           & 0.08 & 0.0784 \\
          \nT  &   \nG           & 0.08 & 0.0792 \\
          \nA  &   \nG           & 0.08 & 0.0798 \\
          \nT  &   \nC           & 0.08 & 0.0795 \\
          \nC  &   \nA           & 0.12 & 0.1202 \\
          \nG  &   \nT           & 0.12 & 0.1196 \\
          \nC  &   \nT           & 0.12 & 0.1198 \\
          \nG  &   \nA           & 0.12 & 0.1193 \\
       \nC\nG  &  \nT\nG         & 0.40 & 0.3978 \\
       \nC\nG  &  \nC\nA         & 0.40 & 0.3978 \\
    \nT\nC\nC  &  \nT\nT\nC      & 0.10 & 0.1033 \\
    \nA\nG\nG  &  \nA\nA\nG      & 0.10 & 0.1033 \\
           \hline
        \end{tabular}
  \end{center}
  \caption{ Results from iterative fitting to a $10^6$ bp sequence
    evolved to have 28.5\% sequence divergence
    with the above mutational motifs. The transitions $CG \to TG$ and $CG \to CA$
    were constrained to be equal (DNA strand symmetry),
    as were the transitions $TCC \to TTC$ and $AGG \to AAG$.
    \label{tab:cpg_results} }
\end{table}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors}
    \end{center}
    \caption{
        Estimated average mutation rates
        for the motifs in the ``Hominid data'' section
        since the common ancestor of humans and chimpanzee.
        Rates are scaled so that the mean time to common ancestor is 0.001 units --
        if we take this to be 6 million years \citep{scally2012insights,langergraber2012generation}
        then rates are in units of mutations per 6,000 years.
        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
        Figure~\ref{sfig:biochem_results_zoom} shows the smaller rates in more detail,
        and Figure~\ref{sfig:biochem_results_relative} shows the rates between categories relative to each other.
        \label{fig:biochem_results}}
\end{figure}

@@

The methods described here provide a way to efficiently calculate full likelihoods,
which in many circumstances provide the optimal basis for inference \citep{neyman1933problem},
fast enough to allow Bayesian inference from Monte Carlo sampling.
The method is fast thanks to its dependence on only local neighborhoods
(so the time does not scale with sequence length once T-mer abundances are counted)
and its reliance on sparse matrices whose structure can be precomputed and cached.
It avoids approximations made by previous methods thanks to the use of T-mers
and explicit quantification of the necessary length of the ``arms'' of the T-mers.
We first discuss possible uses of the method,
followed by challenges (and possible solutions).

@@

\subsection{Acknowledgements}
The initial impetus for this project came from Matt Dean;
along the way substantial encouragement and useful suggestions were provided by
Simon Tavar\'e, Rasmus Nielsen, Graham Coop, Yaniv Brandvain, and Sergey Nuzhdin.
Many thanks are also due to Jessica Crisci for curating the human-chimp data.
The majority of this work was done while PR was at USC.

FAM's work supported by NIH grants R01 AI146028 and U01 AI150747.
Dr.\ Matsen is an Investigator of the Howard Hughes Medical Institute.
